Explore bias sensitivity of Vieira-Silva et al (2019) QMP analysis

### 2022-02-09 notebook/_posts/2022-01-20-vieirasilva2019quan/_cut.Rmd
Revise and render vieirasilva2019quan analysis
Analysis of bias sensitivity of Spearman correlations on QMP profiles
from vieirasilva2019quan.

---
title: "Explore Vieira-Silva et al (2019) QMP profiles"
description: |
author:
  - name: Michael R. McLaren
    url: {}
categories:
  - vieirasilva2019quan
date: 2022-01-20
draft: false
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    dev: svg
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  dpi = 300
)
```

# Setup

## R setup

```{r libraries}
library(here)
# Tools for microbiome data
library(speedyseq)
# Tools for general purpose data manipulation and plotting
library(tidyverse)
library(fs)
# ggplot helpers
library(ggbeeswarm)
library(cowplot)
library(patchwork)
library(scales)
# stats helpers
# library(broom)

library(metacal); packageVersion("metacal")
```

```{r ggplot}
theme_set(theme_cowplot())
```

## Data setup

```{r}
data_path <- here('notebook/_data/vieirasilva2019quan')
dir_create(data_path)
# Supplementary tables from the nature website; contains the metadata
fn <- path(data_path, 'supplementary-data-1.xlsx')
if (!file_exists(fn)) {
  download.file(
    'https://static-content.springer.com/esm/art%3A10.1038%2Fs41564-019-0483-9/MediaObjects/41564_2019_483_MOESM3_ESM.xlsx',
    fn
  )
  file_chmod(fn, '-w')
}
# QMP proiles from the Raes lab website
fn <- path(data_path, 'qmp-matrix-genus.tsv')
if (!file_exists(fn)) {
  download.file('http://raeslab.org/software/QMP2/QMP.matrix.tsv', fn)
  file_chmod(fn, '-w')
}
```

Note, there is an extra value in the BMI column of Table S1 (Excel spreadsheet).

```{r}
otu <- path(data_path, 'qmp-matrix-genus.tsv') %>%
  read.table() %>%
  as('matrix') %>%
  otu_table(taxa_are_rows = FALSE)
sam <- path(data_path, 'supplementary-data-1.xlsx') %>%
  readxl::read_excel(sheet = 'Table S1', range = 'A2:L174', na = "NA")
original_column_names <- names(sam)
ps <- phyloseq(
  otu, 
  sam %>% janitor::clean_names() %>% sample_data
)
stopifnot(setequal(sample_names(ps), sample_names(otu)))
sam <- ps %>% sample_data %>% as_tibble %>% glimpse
```

@@

# Explore

I think that the QMP profiles are such that the sum within a sample should equal the reported cell counts per gram in the sample data; let's check.

```{r}
all.equal(sample_sums(ps), sam$average_faecal_cell_count_cells_g)
```

This small difference may be down to rounding error, since it seems that the entires in the QMP matrix are rounded to integers.

```{r}
otu %>% c %>% {all.equal(., round(.))}
sam$average_faecal_cell_count_cells_g %>% {all.equal(., round(.))}
```

There are `r nsamples(ps)`, with the following disease breakdown:

```{r}
sam %>% count(diagnosis)
```

Can we recover counts?
We can't recover the raw counts without using the CNV information.
But setting that aside for now, let's confirm that we understand the normalization that was used -

```{r}
x <- sam %>%
  add_column(sample_sum = sample_sums(ps))
```
```{r}
x %>% pull(sample_sum) %>% summary
x %>% transmute(sample_sum / average_faecal_cell_count_cells_g) %>% pull(1) %>% summary
```

(we already knew this - the samples sum up to the cell density).

# Reproduce paper analysis

Let's try to reproduce the results from the paper.

> Applying QMP on the combined PSC/IBD/mHC data set, we identified 11 genera as significantly correlat-ing  with  faecal  calprotectin  concentrations  (Spearman  correlation,  n= 149, FDR < 0.05; Fig. 3a and Supplementary Table 6). Stool mois-ture content had 45 associated taxa (Spearman correlation, n= 143, FDR <    0.05;  Supplementary  Table  6).  Combining  both  variables,  only  6  out  of  11  positive(+)  or  negative(−)  calprotectin  associations  remained significant when controlling for faecal water content vari-ation: Anaerostipes(−), Escherichia(+), Fusobacterium(+), Gemmiger(−), Streptococcus(+)  and  Veillonella(+)  (nested  linear  model  on  QMP  rank  abundances,  n= 133,  FDR < 0.05;  Fig. 3a  and  Supplementary  Table  6).

First, let's look for taxa associated with faecal calprotectin concentrations

```{r}
x <- ps %>% as_tibble
```

```{r}
y <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  as_tibble %>%
  with_groups(.otu, nest) %>%
  mutate(
    spearman_test = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman', # exact = FALSE, continuity = TRUE,
        )),
    spearman_test = map(spearman_test, broom::tidy)
  ) %>%
  select(-data) %>%
  unnest(spearman_test) %>%
  mutate(p.bh = p.adjust(p.value, method = 'BH'))
```

```{r}
y %>% arrange(p.bh) %>% print(n=20)
```

```{r}
study_res <- path(data_path, 'supplementary-data-1.xlsx') %>%
  readxl::read_excel(sheet = 'Table S6', na = "NA")
study_res %>%
  select(Genera, starts_with('Calprotectin') & ends_with('QMP'))
y %>% arrange(p.bh) %>% filter(!str_detect(.otu, 'unclassified')) %>%
  select(.otu, estimate, p.value, p.bh) %>%
  print(n=15)
```

These estimates match, but the p values (and their ranking) differs.

figure this out; then try to recapitulate using rank regression.

Note, a decrease FDR-corrected p values can be expected if the FDR correction was done _after_ discarding the OTUs without genus classifications.

can check the code at https://github.com/raeslab/QMP/

```{r}
study_res %>%
  select(Genera, starts_with('Calprotectin') & ends_with('QMP')) %>%
  arrange(4) %>%
  print(n=15)
y %>% 
  arrange(p.bh) %>% 
  filter(!str_detect(.otu, 'unclassified')) %>%
  mutate(
    p.bh = p.adjust(p.value, method = 'BH'),
    across(starts_with('p.'), round, digits = 4)
  ) %>%
  select(.otu, estimate, p.value, p.bh) %>%
  print(n=15)
```

ahhh - ok their's is sorted differently, which makes sense since it has many tests in one table. and the p values need to be rounded to 4 digits. then it seems like we're good.

next:

- also do for moisture, and perhaps RMP, to check these results as well
- confirm I can get similar result with linear model on ranks
<!--  -->

visually inspect

```{r}
top_hits <- y %>% 
  filter(!str_detect(.otu, 'unclassified')) %>%
  slice_min(p.bh, n = 9)
x <- ps %>%
  filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  prune_taxa(top_hits$.otu, .) %>%
  as_tibble

x %>%
  mutate(across(.otu, factor, levels = top_hits$.otu)) %>%
  with_groups(.otu, mutate,
    across(c(faecal_calprotectin_mg_g, .abundance), rank),
  ) %>%
  # mutate(across(c(faecal_calprotectin_mg_g, .abundance), rank, ties.method = 'average')) %>%
  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
  facet_wrap(~.otu) +
  geom_point() +
  stat_smooth(method = 'lm')
```

It seems like zeros or below-detection-limit values are largely responsible for the observed results.

```{r}
x %>%
  ggplot(aes(faecal_calprotectin_mg_g, .abundance)) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~.otu) +
  geom_point()
```

@@

### Check if patterns are driven by presence/absence

```{r}
y <- ps %>%
  # filter_sample_data(!is.na(faecal_calprotectin_mg_g)) %>%
  # prune_taxa(str_subset(taxa_names(.), 'unclassified', negate = TRUE), .) %>%
  as_tibble %>%
  with_groups(.sample, mutate, 
    proportion = close_elts(.abundance),
    present = (.abundance > 0) * 1
  ) %>%
  with_groups(.otu, nest) %>%
  mutate(
    spearman = map(data, 
      ~cor.test(data = .x,
        ~ .abundance + faecal_calprotectin_mg_g, 
        method = 'spearman')),
    spearman_pa = map(data, 
      ~cor.test(data = .x,
        ~ present + faecal_calprotectin_mg_g, 
        method = 'spearman')),
  ) %>%
  select(-data) %>%
  pivot_longer(contains('spearman'), names_to = 'test', values_to = 'fit') %>%
  mutate(across(fit, map, broom::tidy)) %>%
  unnest(fit) %>%
  with_groups(test, mutate, p.bh = p.adjust(p.value, method = 'BH'))
```

```{r}
y.w <- y %>%
  select(.otu, test, estimate, p.bh) %>%
  pivot_wider(names_from = c(test), values_from = c(estimate, p.bh), 
    names_glue = '{test}_{.value}', names_sort = TRUE) %>%
  select(.otu, sort(tidyselect::peek_vars())) %>%
  mutate(across(where(is.numeric), round, 4)) %>%
  arrange(spearman_p.bh)
```

```{r}
y.w %>%
  ggplot(aes(spearman_estimate, spearman_pa_estimate)) +
  geom_point()
```

For some taxa (like Veillonella, EscherichiaShigella) the two estimates are very similar;for others like Streptococcus they are quite different.
I suspect that simulated bias will have a larger impact on Streptococcus.

# Bias sensitivity analysis

```{r, eval = F}
# Perturb by bias vector - treat ps as actual
ps.per <- perturb(ps, bias, norm = 'keep')
# Calibrate by bias vector - treat ps as observed
ps.cal <- calibrate(ps, bias, norm = 'keep')
```

@@

```{r}
my_stat <- function(ps, otu) {
  data <- ps %>%
    prune_taxa(otu, .) %>%
    as_tibble
  spearman_test <- cor.test(data = data,
    ~ .abundance + faecal_calprotectin_mg_g, 
    method = 'spearman', exact = FALSE
  )
  spearman_test %>% broom::tidy()
}

@@

Example where the bias sensitivity analysis shows little effect of varying the hypothesized bias.

We can probably understand this finding because the original result is driven by the presence/absence pattern, and we have treated zeros as absolute, so that presence/absence is maintained.
If we relaxed this assumption, then I suspect that we would see much more variation in the estimates.
Much of this variation, however, would come from relaxing our assumption about the zeros, rather than relaxing our assumption that there is no bias.

Challenge with this data is that we don't have the read counts, so don't know how to replace zeros with resonable values.

## saved plotting stuff

x %>%
  ggplot(aes(x = mean_efficiency, fill = faecal_calprotectin_mg_g > 50)) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
x %>%
  ggplot(aes(x = faecal_calprotectin_mg_g, y = mean_efficiency, color = diagnosis)) +
  scale_y_log10() +
  scale_x_log10() +
  geom_point() +
  stat_smooth(aes(group = 1), method = 'lm')
```

Now, redo the analysis with and without bias

what assumptions to make about zeros?

>>>>

@@

@@

# saved plotting stuff

x %>%
  ggplot(aes(x = mean_efficiency, fill = faecal_calprotectin> 50)) +
  scale_x_log10() +
  geom_histogram()
```

```{r}
x %>%
  ggplot(aes(x = faecal_calprotectin, y = mean_efficiency, color = diagnosis)) +
  scale_y_log10() +
  scale_x_log10() +
  geom_point() +
  stat_smooth(aes(group = 1), method = 'lm')
```

# Checks and notes

```{r}
# check na handling
x = tibble(a = c(1,2,3,NA,4,5), b = c(6,5,4,3,2,NA))
cor.test(~a + b, data = x, method = 'spearman') %>% broom::tidy()
cor.test(~a + b, data = x %>% drop_na, method = 'spearman') %>% broom::tidy()
```

## Original bias sens analysis

@@

```{r}
my_stat <- function(ps, otu) {
  data <- ps %>%
    prune_taxa(otu, .) %>%
    as_tibble
  spearman_test <- cor.test(data = data,
    ~ .abundance + faecal_calprotectin, 
    method = 'spearman', exact = FALSE
  )
  spearman_test %>% broom::tidy()
}

@@
